A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-11 paper.tex
Figure 6 narrative

The problem of reconstructing ARGs for samples of recombining sequences has been of
interest since the ARG was first defined. Early methods focused on finding \emph{parsimonious} ARGs,
i.e.\ those with a minimal number of recombination events \citep{hein1990reconstructing}.
Two main approaches emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}.
The former starts with a data matrix and reduces it to an empty matrix through row and column operations
corresponding to coalescence, mutation, and recombination events, which construct
an ARG from the bottom up. The latter begins from an initial local tree at a single focal site. Moving the
focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
whenever a recombination is traversed. Methods based the backwards-in-time approach are described by
\citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
while \citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome methods.
\citet{rasmussen2022espalier} focuses on parsimonious fusion of local trees into an ARG, while
\citet{camara2016inference} is based on topological data analysis.

Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
so parsimony-based methods resort to heuristics and are limited to analysing hundreds of sequences.
Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
rather than parsimonious ARGs
\citep{minichiello2006mapping, parida2008estimating, kelleher2019inferring,  speidel2019method, zhang2021biobank}.
The latter three methods, as well as the parsimony-based method of \citet{rasmussen2022espalier},
can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.

@@

>>>>

\subsection*{ARG reconstruction}

The problem of reconstructing ARGs for samples of recombining sequences has been of
interest since the ARG was first defined. Early methods focused on finding \emph{parsimonious} ARGs,
i.e.\ those with a minimal number of recombination events \citep{hein1990reconstructing}.
Two main approaches emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}.
The former starts with a data matrix and reduces it to an empty matrix through row and column operations
corresponding to coalescence, mutation, and recombination events, which construct
an ARG from the bottom up. The latter begins from an initial local tree at a single focal site. Moving the
focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
whenever a recombination is encountered. Methods based the backwards-in-time approach are described by
\citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
while \citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome methods.
\citet{rasmussen2022espalier} focuses on parsimonious fusion of local trees into an ARG, while
\citet{camara2016inference} is based on topological data analysis.

Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
so parsimony-based methods resort to heuristics and are limited to analysing at most hundreds of sequences.
Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
rather than parsimonious ARGs
\citep{minichiello2006mapping, parida2008estimating, kelleher2019inferring,  speidel2019method, zhang2021biobank}.
The latter three methods, as well as the parsimony-based method of \citet{rasmussen2022espalier},
can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.

@@

\subsection*{Differences in reconstructed ARGs}

There are a vast number of ARGs compatible with a given dataset, so ARG reconstruction tools generally produce different outputs for the same input. Figure \ref{fig-inferred-args} shows examples of the ARG topologies produced by four of the tools mentioned above, applied to the classical dataset of 11 \emph{Drosophila melanogaster} sequences of \citet{kreitman1983nucleotide}.

\texttt{KwARG} \citep{ignatieva2021kwarg} produces a parsimonious ARG with seven recombination breakpoints, which has been shown to be the true minimum required to explain the dataset when disallowing recurrent mutation \citep{song2003parsimonious}. An ARG drawn from the posterior sample produced by \texttt{ARGweaver} has 17 breakpoints. Both tools explicitly record the sequence undergoing each recombination event, and the precise location of the breakpoint along the genome. \texttt{Relate} \citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} instead construct a sequence of correlated local trees, but without specifying the precise positions of recombination events (i.e.\ reconstructing the local trees without explicitly specifying the SPR operations required to transition between them). The number of inferred trees also depends on the parameter settings; all four tools can allow for the presence of sequencing errors and recurrent mutations, reducing the number of inferred local trees (in this case, based on the input mutation and recombination rate \texttt{Relate} hence posits only two recombination breakpoints).

There is some consistency among the four inferred ARGs (such as the samples FR-F, WA-F and Af-F being grouped into the same clade), but also considerable differences in the order and number of events. Note that all four ARGs can be represented in the succinct tree sequence format, with the ARGs produced by \texttt{KwARG} and \texttt{ARGweaver} containing more nodes to record the recombination events (corresponding to Figure 4 stage `B'), and the ARGs output by \texttt{tsinfer} and \texttt{Relate} being, in essence, simplified (Figure 4 stage `D').
